---
title: "Homework 3 Checker"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(learnrhash)
library(tidyverse)
library(gradethis)
library(openintro)
tutorial_options(
  # use gradethis for checking
  exercise.checker = gradethis::grade_learnr
  )
knitr::opts_chunk$set(echo = FALSE)
set.seed(113877)
```


## Getting started

To check Question 1c from the MGF examples (part of HW 3) we will need to generate random numbers from a specific normal distribution. How can we find out what parameters the `rnorm` function expects us to specify?

```{r getHelp, exercise = TRUE}
___(rnorm)
```

```{r getHelp-solution}
help(rnorm)
```

```{r getHelp-check}
# check code
gradethis::grade_code()
```


```{r whichspread}
question("Which measure of spread does the function rnorm expect?",
    answer("variance"),
    answer("standard deviation", correct = T),
    answer("doesn't matter"),
    allow_retry = TRUE,
    random_answer_order = F
  )
```

## Checking Question 1c from the MGF examples (part of HW 3) 

```{r checkSol, exercise = TRUE}
n <- 100000 ## number of draws from the distribution, just pick something large

mu <- ___ ## mean parameter
sigma_sq <- ___ ## variance parameter

test <- rnorm(n = ___, mean = ___, sd = sqrt(___)) 

mean(test^2 + exp(0.5*test))
```

```{r checkSol-solution}
n <- 100000 ## number of draws from the distribution, just pick something large

mu <- 5 ## mean parameter
sigma_sq <- 4 ## variance parameter

test <- rnorm(n = n, mean = mu, sd = sqrt(sigma_sq)) 

mean(test^2 + exp(0.5*test))
```

```{r checkSol-check}
# check code
gradethis::grade_code()
```

Now write your derived expectation (as a formula) and check if the value is close to the one we observed empirically.

```{r eqnSolution-setup}
mu <- 5 ## mean parameter
sigma_sq <- 4 ## variance parameter
```

```{r eqnSolution, exercise = TRUE}
true_expectation <- ___

true_expectation ## print out answer
```





