---
title: "Your First Homework Helper"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(learnrhash)
library(tidyverse)
library(gradethis)
tutorial_options(
  # use gradethis for checking
  exercise.checker = gradethis::grade_learnr
  )
knitr::opts_chunk$set(echo = FALSE)
set.seed(113877)
```

## Getting started

This is a `learnr` tutorial. It allows you to edit and run R code with some automatic feedback about whether you are on the right track. You can edit the code and click "Run Code" to see what happens.

```{r test-run, exercise = TRUE}
(2+5)*3-11

5^3
```

For *assignment purposes* in R, `=` and `<-` can be used interchangeably.  The arrow is the preferred notation.

```{r test-run2, exercise = TRUE}
y <- 3
x <- 5 - 9*y

x
```

To learn what a function does and what paramters it expects, you can consult the help pages.

```{r get-help, exercise = TRUE}
help(exp)
```

## Checking HW 2, Question 2

Suppose that $X_1 \sim Binomial(n=20, p = 1/5)$ and $X_2 \sim Poisson(\lambda = 3)$. Further assume that $X_1$ and $X_2$ are independent. Let $Y = 3X_1 + 2X_2$.

a.) Find $E[Y]$.

b.) Find $V(Y)$.

```{r simSolution, exercise = TRUE}
sample_size <- 10000 ## number of draws from the distribution, just pick something large
n <- ___ ## number of trials
p <- ___ ## probability of success, write as a fraction
lambda <- ___ ## Poisson parameter

X1 <- rbinom(n = sample_size, size = n, p = p) ## draw from the Binomial distribution
X2 <- rpois(n = sample_size, lambda = lambda) ## draw from the Poisson distriubtion

Y <- 3*X1 + 2*X2

mean(Y) ## empirical mean
var(Y) ## empirical variance
```

```{r simSolution-solution}
sample_size <- 10000 ## number of draws from the distribution
n <- 20 ## number of trials
p <- 1/5 ## probability of success, write as a fraction
lambda <- 3 ## Poisson parameter

X1 <- rbinom(n = sample_size, size = n, p = p) ## draw from the binomial distribution
X2 <- rpois(n = sample_size, lambda = lambda) ## draw from the Poisson distriubtion

Y <- 3*X1 + 2*X2

mean(Y) ## empirical mean
var(Y) ## empirical variance
```

```{r simSolution-check}
# check code
gradethis::grade_code()
```


Now write your mean and variance in terms of the parameters $n$, $p$, and $\lambda$ (as a formula) and check if the theoretical mean and variance are close to the ones we observed empirically.

```{r eqnSolution-setup}
n <- 20 ## number of trials
p <- 1/5 ## probability of success
lambda <- 3 ## Poisson parameter

```

```{r eqnSolution, exercise = TRUE}
true_mean <- ___
true_var <- ___

true_mean ## print out answer
true_var ## print out answer
```
