---
title: "Guess the Color"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
library(learnrhash)
library(tidyverse)
library(gradethis)
tutorial_options(
  # use gradethis for checking
  exercise.checker = gradethis::grade_learnr
  )
knitr::opts_chunk$set(echo = FALSE)
set.seed(113877)

generateColor <- function(num){
  samp = rbinom(num, 1, prob = 0.75)
  show = ifelse(samp == 1, "green","red")
  return(show)
}
```

## One Observation

I have a function that returns either "red" or "green". We don't know anything else about it. 

```{r, one-light, exercise = T}
generateColor(1)
```

**How can we predict what the next color is?**

## More Observations

It might be helpful to see more observations to better understand the function `generateColor`.

```{r, more-obs, exercise = T}
generateColor(10)
```


## Your Strategy

So we've learned that `generateColor` produces "green" about 75% of the time and "red" about 25% of the time. 

```{r, student-strategy, exercise = T}


```

**What is your strategy of predicting what the next color is?**

**How would you assess whether or not your strategy is "better" than someone else's?**

**What does it mean to be "better?"**


## Strategy 1

What if we just stuck with the most likely outcome each time and always picked "green"?

```{r, strat1, exercise = T}
obs <- generateColor(1000)
guess <- rep("green", 1000)
```

**What is the probability that we correctly guess the next color using this approach?**

**What is the expected number of correct guesses out of 50 using this approach?**

## Strategy 1 Performance - Simulation

We can double check our calculations by using simulation. 

```{r, strat1-perf, exercise = T}
obs <- generateColor(50)
guess <- rep("green", 50)

## number of correct guesses
sum(obs == guess)

## proportion of correct guesses
sum(obs == guess)/50
```

**What's the relationship between this proportion and the probability we are investigating?**

**How could I make this empirical estimate more accurate?**

```{r, strat1-perf2, exercise = T}

```

## Strategy 2

```{r, strat2, exercise = T}
obs <- generateColor(1000)

guess <- rbinom(1000, 1, prob = 0.75)
guess <- ifelse(guess == 1, "green", "red")
```


**What is the probability that we correctly guess the next color using this approach?**

**What is the expected number of correct guesses out of 50 using this approach?**


## Strategy 2 Performance - Simulation

We can double check our calculations by using simulation. 

```{r, strat2-perf, exercise = T}
obs <- generateColor(50)

guess <- rbinom(50, 1, prob = 0.75)
guess <- ifelse(guess == 1, "green", "red")

## number of correct guesses
sum(obs == guess)

## proportion of correct guesses
sum(obs == guess)/50
```

## Does anyone have a different strategy?

```{r, your-strat, exercise = T}
obs <- generateColor(1000)
guess <-

## number of correct guesses
sum(obs == guess)

## proportion of correct guesses
sum(obs == guess)/50
```

## Variance

We have a random variable that represents whether or not we get the next guess correct. What is the variance of that random variable in both scenarios?

```{r, var, exercise = T}
obs <- generateColor(1000)

guess1 <- rep("green", 1000)

guess2 <- rbinom(1000, 1, prob = 0.75)
guess2 <- ifelse(guess2 == 1, "green", "red")

var(obs == guess1)
var(obs == guess2)
```



* * *

This lab was adapted and learnr-ified by S. Stoudt from the Can You "Beat" Randomness (Zieffler and Catalysts for Change 2013) activity and the adaptation of this activity (Green and Blankenship 2015).


