---
title: "Probability Simulations"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(gradethis)
knitr::opts_chunk$set(echo = FALSE)
```

**Note: This tutorial is only minorly adapted based on a tutorial written and shared by Anthony Scotina.**

## Introduction

When I Google the term "*simulation*", the first result that pops up is the *Oxford Languages* definition:

> sim·u·la·tion (noun): Imitation of a situation or process.

A major goal of probability and statistical inference is to translate data into knowledge and understanding of the real world. And while probabilistic experiments can be *costly*, *tedious*, or downright *impractical* to perform in real life (imagine tossing a coin 1,000 times!), **Monte-Carlo simulations** can help us model these experiments and *imitate* the results that we might expect to see in a real life scenario. 

Simulations give us a *ton* of power in a "theoretical" class like *Probability*. If we have a solid understanding of the experimental setting and *how* to actually set up the simulation, we can answer many probability questions that could otherwise be quite challenging to evaluate by hand. 

For this lab, we will focus on two core **base R** functions that are used to simulate data:

- `sample()`: takes a sample of elements from a vector, with or without replacement

- `replicate()`: repeats an experiment many times, and (usually) returns a vector as the result

We'll also leverage some R tools we've seen already: **vectors**, **indexing**, **logicals**, and more...

Let's get started with some simulations!

## Simple Simulations with `sample()`

The `sample()` function is a simple, yet powerful tool. If we run `?sample`, we can see more information, but it boils down to the following syntax:

```{r, eval = FALSE}
sample(x, size, replace = FALSE, prob = NULL)
```

- `x` is the vector of all *objects* from which we are sampling. 

- `size` is the number of objects we would like to select from `x`. If I would like to choose **5** numbers at random, I would enter `size = 5`. 

- `replace` can take one of two values: `TRUE` or `FALSE` - these are case-sensitive, and R handles these two **logical** values rather carefully.
    - If `replace = TRUE`, we will sample from `x`, `size` times, **with replacement**. In other words, if we select *five numbers* between 1 and 100 *with replacement*, we could potentially choose any of those numbers *more than once*. 
    - If `replace = FALSE`, sampling is performed **without replacement**. This is the **default** setting. 
    
- `prob` is a *vector* of **probability weights**. This assigns a probability to each element in your vector, `x`. 
    - If we would like to simulate flipping a *biased coin*, the vector `c("H", "T")` might have the probability weights of `prob = c(0.75, 0.25)`. This just means that our biased coin has a 75% probability of landing *Heads*. 
    - `prob = NULL` by **default**, which means it assumes each element from `x` has an equally-likely chance of being sampled. 
    
###

We can specify the vector `x` in a number of ways. For example...

- `c("H", "T")` is one way to specify the two possible outcomes from a *single coin flip*. 

- `1:100`, `seq(from = 1, to = 100, by = 1)` or `c(1, 2, 3, 4, ..., 99, 100)` are three different ways to specify a vector of the numbers 1 through 100. 
    - I personally prefer `1:100` - if you use `c(...)`, you would actually have to type out *every single number* from 1 through 100. I got lazy and added the `...` instead! 

- There are several ways we could specify a vector of 52 playing cards. 
    - If we don't care about the suit or number, we could just sample from `1:52`, treating each card uniquely. 
    - If we *did* care about the suit of the card, we could type something like:
    ```{r, eval = FALSE}
    c(rep("Hearts", 13), rep("Diamonds", 13), rep("Spades", 13), rep("Clubs", 13))
    ```
    
> 🚨**NOTE**🚨: The `sample()` will never return the same output twice, unless you run `set.seed(338)` (or with any other number) just before `sample()`. This will control R's built-in *random number generator*.
    
### Exercise 1 

There are **8** people in this class, including Prof. Stoudt. Let's suppose each person in the class is assigned a different number from 1 through 8. Using `sample()`, randomly select four numbers from this list, **with replacement**. 

```{r, ex1, exercise = TRUE}

```

```{r, ex1-solution}
sample(1:8, size = 4, replace = FALSE)
```

### Exercise 2

Simulate flipping a coin ten times. In this context, why is it essential to sample *with replacement*?

```{r, ex2, exercise = TRUE}

```

```{r, ex2-solution}
sample(c("H", "T"), size = 10, replace = TRUE)
```

### Exercise 3

Suppose we would like to flip a six-sided die, but the sides, *y*, are biased, such that:

```{r, echo = FALSE}
data.frame(y = "Prob(y)", one = 0.05, two = 0.25, three = 0.1, four = 0.1, five = 0.4, six = 0.1) 
```

Simulate rolling this *biased* die 20 times. 

```{r, ex3, exercise = TRUE}

```

```{r, ex3-solution}
sample(1:6, size = 20, replace = TRUE, 
       prob = c(0.05, 0.25, 0.1, 0.1, 0.4, 0.1))
```


## Estimating Probabilities

While using `sample()` on its own is nice (I guess?), we haven't actually used it to calculate **probabilities** yet! What we've done so far is use `sample()` to simulate an experiment *once*. But probability (at least when using the [frequentist definition](https://en.wikipedia.org/wiki/Frequentist_probability)) involves repeating a random experiment a large number of times. When we say a fair coin will land "Heads" 50% of the time, we really mean it would land "Heads" 50% of the time **if we tossed the coin over and over again**. 

Let's further explore this *not drawn out at all* coin flipping example. 

To sample a *single coin toss*, we could use `sample()` as such:

```{r, coin_toss1, exercise = TRUE, exercise.eval = FALSE}
sample(c("H", "T"), size = 1)
```

- Note that the `replace` option is irrelevant here since we're only flipping the coin once.

Now run that same line *over and over* again. You'll get "H" sometimes, and "T" the other times. You might get three "H" in a row, followed by five "T" in a row. Each toss is equally likely, and *independent* from the previous toss. 

### 

To repeat this experiment, say, **10,000 times**, we'll use the `size` option, and save this as `coin_tosses` (because no one wants to see 10,000 values listed out). Don't forget `replace = TRUE` here, since we're repeating this experiment more than once!

```{r, prepare-coins, exercise = TRUE, exercise.eval = FALSE}
set.seed(338) # Change to your favorite number!
coin_tosses = sample(c("H", "T"), size = 10000, replace = TRUE)
```

Next we'll want to check how many elements from the `coin_tosses` vector have the value "H". We can do this by first converting `coin_tosses` to a **logical** vector, and then counting up the number of `TRUE` values:

```{r, coins_sum, exercise = TRUE, exercise.eval = FALSE, exercise.setup = "prepare-coins"}
sum(coin_tosses == "H")
```

You could optionally run `coin_tosses == "H"` to see what exactly we are summing up. It's a bunch of `TRUE` and `FALSE` elements, which hold values of **1** and **0** in R, respectively. I simulated 4,960 heads out of 10,000 coin tosses, *but this still isn't a probability*! Remember, a probability is a value between **0 and 1**. 

###

You *could* just divide this by 10,000, but there is an easier way that we'll use a bunch this semester: Simply take the `mean()` of the logical vector of 1s and 0s to get the *proportion* of 1s (i.e., `TRUE` values):

```{r, coins_mean, exercise = TRUE, exercise.eval = FALSE, exercise.setup = "prepare-coins"}
mean(coin_tosses == "H")
```

We *know* that the probability is actually **0.50**. But I *estimated* a probability of 0.496. If I simulated even more coin flips, I'd probably get even close to that true probability of 0.50; this fact is summarized in one of the most important theorems in probability theory: [**The Law of Large Numbers**](https://en.wikipedia.org/wiki/Law_of_large_numbers). 

### Exercise 4

Repeat this simulation, but for tossing a fair, six-sided die. Estimate the probability of rolling a **5**. How close is this to the "true" probability?

```{r, ex4, exercise = TRUE}

```

```{r, ex4-solution}
die_rolls = sample(1:6, size = 10000, replace = TRUE)
mean(die_rolls == 5)
```



## Larger Simulations with `replicate()`

We'll often be faced with more complicated examples that don't necessary involve repeating a *single trial* (i.e., rolling a die *once*) over and over again. The `replicate()` function is an example of an **implicit loop** in R, and it will help us with such calculations. 

Here's the setting for this example: Suppose I am working with a standard deck of 52 cards, and I choose two cards at random from the deck. Jacks, queens, and kings each have a value of 10, and aces have a value of 11. Each of the number cards has the value of that number. 

> What is the probability that the **sum** of my two cards is exactly 21?

###

First let's work by simulating this experiment **once**, as we did before. To do this, we need to specify the vector `x`, which is the set of all cards from which we're sampling. But we care about each card's **value** here, *not* the suit. So we can just sample from the following vector twice, *without replacement*:

```{r, prepare-cards}
# Repeat this vector FOUR times, once for each suit
cards = rep(c(2,3,4,5,6,7,8,9,10,10,10,10,11), 4)
```

(**Note**: There isn't actually a "1" card in a standard deck, and we're treating the "Ace" card as having a value of **11** here.)

- Visit [this page](https://en.wikipedia.org/wiki/Standard_52-card_deck) for more about a standard deck of playing cards!

Using the `cards` vector, let's choose two cards, and add up their values:

```{r, cards1, exercise = TRUE, exercise.eval = FALSE, exercise.setup = "prepare-cards"}
set.seed(338) # Change to your favorite number!
card_draw = sample(cards, size = 2, replace = FALSE)
card_draw

sum(card_draw)

sum(card_draw) == 21
```

The sum of my two cards is **19**, which isn't quite 21... I guess I lost. But if I repeated this *over and over again*, how many times *would* I win?

###

The `replicate()` function can help us calculate more complicated probabilities, and it takes two values:

```{r, eval = FALSE}
replicate(n, expr)
```

- `n` is the number of times I want to repeat the single trial. I usually set this to 10,000. 

- `expr` is the *block of code* we wish to repeat `n` times. 

For example, these two blocks of code are equivalent:

- The `sample()` 10,000 times approach:

    ```{r}
    set.seed(338) # Change to your favorite number!
    coin_tosses = sample(c("H", "T"), size = 20, replace = TRUE)

    mean(coin_tosses == "H")
    ```

- The `replicate()` 10,000 times approach:

    ```{r}
    set.seed(338)
    heads = replicate(20, {
      coin_toss = sample(c("H", "T"), size = 1)
      coin_toss == "H"
    })
    heads
    mean(heads)
    ```

> 🚨**NOTE**🚨: In `replicate()`, the curly brackets `{ }` are required when you are replicating more than one R expression!

Of course, the `replicate()` approach is more complicated here, but we applied it to a relatively simple calculation! It would make more sense to use the simple approach!

###

For our *blackjack/21* example, `replicate()` is the way to go. We just want to replicate the following code a bunch of times:

```{r, cards2, exercise = TRUE, exercise.eval = FALSE, exercise.setup = "prepare-cards"}
card_draw = sample(cards, size = 2, replace = FALSE)

sum(card_draw) == 21
```

The output from `replicate()` will be a vector of 10,000 (or whatever you use for the `n` option) `TRUE` and `FALSE` values. Thus, to find the probability, we simply need to take the `mean()` of this vector. 

```{r, cards3, exercise = TRUE, exercise.eval = FALSE, exercise.setup = "prepare-cards"}
sum_21 = replicate(10000, {
  card_draw = sample(cards, size = 2, replace = FALSE)
  sum(card_draw) == 21
})
mean(sum_21)
```

### The `replicate()` Workflow

In MATH/STAT 338, we'll encounter many probability examples that we can derive *by hand* and/or *estimate through simulation*. Though there will be many cases where the simulation-based approach will provide a simpler and more intuitive way to calculating probabilities. 

For the especially complicated simulation-based exercises that you'll encounter, it is a good idea to follow a **workflow**, or a set of steps that you can repeat for similar exercises. For example, I recommend you follow this workflow from [Speegle and Clair (2021)](https://mathstat.slu.edu/~speegle/_book/preface.html) when using `replicate()` to estimate probabilities:

1. Write code that performs the experiment a single time. For example, if we are estimating the **probability of obtaining at least 12 heads in 20 coin tosses**, a single experiment would be *one set of 20 coin tosses*:

    ```{r}
    coin_tosses = sample(c("H", "T"), size = 20, replace = TRUE)
    sum(coin_tosses == "H") >= 12
    ```

2. Replicate the experiment a small number of times and **check** the results:

    ```{r, eval = FALSE}
    replicate(100, {
      EXPERIMENT GOES HERE
      ...
    })
    ```

3. Replicate the experiment a large number of times and **store** the results:

    ```{r, eval = FALSE}
    event = replicate(10000, {
      EXPERIMENT GOES HERE
      ...
    })
    ```

4. Estimate the probability with `mean(event)`. Note that `event` will *always* be a **logical vector**, so you'd be taking the mean of a bunch of 1s and 0s. 

### Exercise 5 

Try to simulate each of the three events stated below using `replicate()`. 

Isaac Newton was consulted about the following problem by Samuel Pepys, who wanted the information for gambling purposes. Which of the following events has the highest probability?

- *A*: At least one 6 appears when 6 fair dice are rolled. 

- *B*: At least two 6's appear when 12 fair dice are rolled. 

- *C*: At least three 6's appear when 18 fair dice are rolled. 

```{r, ex5, exercise = TRUE}

```

```{r, ex5-solution}
A = replicate(10000, {
  dice_roll = sample(1:6, size = 6, replace = TRUE)
  sum(dice_roll == 6) >= 1
})
mean(A)

B = replicate(10000, {
  dice_roll = sample(1:6, size = 12, replace = TRUE)
  sum(dice_roll == 6) >= 2
})
mean(B)

C = replicate(10000, {
  dice_roll = sample(1:6, size = 18, replace = TRUE)
  sum(dice_roll == 6) >= 3
})
mean(C)
```



## Conclusion

That's it for now! With `sample()` and `replicate()`, you'll have a lot of power to *estimate* sometimes complicated probabilities via **Monte-Carlo simulation**. 

Thanks again to Anthony Scotina for sharing this tutorial with us!
    










    







