---
title: "Coding Training 2: Structures and Subsetting"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(learnrhash)
library(tidyverse)
library(gradethis)
tutorial_options(
  # use gradethis for checking
  exercise.checker = gradethis::grade_learnr
  )
knitr::opts_chunk$set(echo = FALSE)
```

## Code Annotation

As you learn about R code you will encounter code chunks that you don't fully understand at first. A good way to approach this scenario is read through the code and going through the following annotation steps.

  
1. Do an initial reading of each code chunk. Note any questions that pop into your head. If you do this inline, use a hashtag (#) to add a comment.
    
2. Identify any functions you do not understand and look them up using the help function `help(function_name)` or the "Help" tab in the bottom right pane. Make sure you understand the input paramters and the form of the output. 
    
3. Identify any overall structure. Is a subtask repeated multiple times? How many times?
      
4. For each object, predict their size. Check your intuition using `dim` or `length`. You may also want to use `class` to identify the type of an object.
    
5. Write a one to two sentence description of what the code chunk does that someone without a coding background could understand.

You may want to use this for a subset of code chunks in this tutorial.


## Data Management

### Vectors

The `c()` function *concatenates* items (of homogenous type) into a vector; for example, `myvector = c(1, 3, 6, 9)` creates a vector that has four elements. To learn about how many elements in the vector there are, you can use the function `length`.

```{r myvec exercise = T}
myvector <- c(1, 3, 6, 9)
myvector

length(myvector)
```

To access a particular element, you can use square brackets.

```{r myvec2 exercise = T}
myvector <- c(1, 3, 6, 9)

myvector[2]
myvector[c(2,3)]
```

Use `c()` to create a vector called `myvector` that contains all the perfect squares (1, 4, 9, 16, $\dots$) less than or equal to 100.

```{r makeVec, exercise = TRUE}

```

```{r makeVec-solution}
myvector <- c(1, 4, 9, 16, 25, 36, 49, 64, 81, 100)
```

```{r makeVec-check}
# check code
gradethis::grade_code()
```

### Matrices

What are the dimensions of the following matrix? Use `dim` to find out.

```{r makeMatrix, exercise = T}
mymatrix <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 3)

dim(___)
```

```{r makeMatrix-solution}
mymatrix <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 3)

dim(mymatrix)
```

```{r makeMatrix-check}
# check code
gradethis::grade_code()
```


We can access specific rows of a matrix with `[,]` - for example, `mymatrix[2,]` will present the second row of the matrix, and `mymatrix[c(2,3),]` will yield a new matrix with the second and third rows of the original only.


```{r makeMatrix2, exercise = T}
mymatrix <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 3)

mymatrix[c(2,3),]
```

Assign the first row of `mymatrix` to a variable named `row1`. Use the function `summary()` to  find the minimum and maximum of this row.

```{r matrixWork, exercise = T}
mymatrix <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 3)

___ <- mymatrix[___]
summary(___)
```

```{r matrixWork-solution}
mymatrix <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 3)

row1 <- mymatrix[1,]
summary(row1)
```

```{r matrixWork-check}
# check code
gradethis::grade_code()
```

To get a single element in a matrix, we use row and column numbers, such as `mymatrix[r,c]`. This will also extract a submatrix if multiple entries are used for each.

```{r matrixWork2, exercise = T}
mymatrix <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 3)

mymatrix[1,2]
```

Display the submatrix that contains the first two rows and second column of `mymatrix`. 


```{r matrixWork3, exercise = T}
mymatrix <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 3)

mymatrix[___, ___]
```

```{r matrixWork3-solution}
mymatrix <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 3)

mymatrix[c(1,2), 2]
```

```{r matrixWork3-check}
# check code
gradethis::grade_code()
```

## Logicals


We can also ask questions about our data/variables. The `==` sign is a True/False test of equality. e.g. `row2 == 1` returns `TRUE` for every element in row2 that is equal to 1, and `FALSE` for elements different than 1. We can do similar things with the symbols `<`, `<=`, `>`, `>=`. 

```{r log exercise = T}
row2 <- c(1, 2, 3, 1)

row2 == 1
```


You can also combine two conditions with the AND (`&`) and OR (`|`) operators.

```{r log2 exercise = T}
row2 <- c(1, 2, 3, 1)
col3 <- c(1, 1, 3, 2)

row2 >= 100 & col3 == 1

row2 == 1 | col3 < 100

sum(row2 == 1)

which(row2 == 1)
```


```{r quiz}
quiz(
  question("The sum line does the following:",
    answer("counts how many elements are in row2"),
    answer("counts how many elements are not equal to 1 in row2"),
    answer("counts how many 1s there are in row2", correct = TRUE),
    answer("sums the elements in row2"),allow_retry = TRUE,
    random_answer_order = T
  ),
  question("The which line does the following:",
    answer("sums the elements in row2 that are equal to 1"),
    answer("counts how many elements are equal to 1 in row2"),
    answer("tells the positions of the elements in row2 that are equal to 1", correct = TRUE),
    answer("tells the positions of the elements in row2 that are not equal to 1"),allow_retry = TRUE,
    random_answer_order = T
  )
)
```


We can use these conditions to select subsets.

Return the values in `col1` where the corresponding values of `col3` equal 1.

```{r subset1, exercise = T}
col1 <- c(4, 5, 6, 4)
col3 <- c(1, 1, 3, 2)

col1[___]
```

```{r subset1-solution}
col1 <- c(4, 5, 6, 4)
col3 <- c(1, 1, 3, 2)

col1[col3 == 1]
```

```{r subset1-check}
# check code
gradethis::grade_code()
```


We can analyze these subsets as well. Compute the average of only the `col3` values for whom the corresponding `col1` values are 4.

```{r subset2, exercise = T}
col1 <- c(4, 5, 6, 4)
col3 <- c(1, 1, 3, 2)

mean(___)
```

```{r subset2-solution}
col1 <- c(4, 5, 6, 4)
col3 <- c(1, 1, 3, 2)

mean(col3[col1 == 4])
```

```{r subset2-check}
# check code
gradethis::grade_code()
```



Find the minimum of the values in col1 whose col3 values are less than 15.

```{r subset3, exercise = T}
col1 <- c(4, 5, 6, 4)
col3 <- c(10, 10, 30, 2)

min(___)
```

```{r subset3-solution}
col1 <- c(4, 5, 6, 4)
col3 <- c(10, 10, 30, 2)

min(col1[col3 < 15])
```

```{r subset3-check}
# check code
gradethis::grade_code()
```


*Note:* when your data values are text, i.e. letters or words, to use the == sign or any of these other logical operators, you need to put quotes around the text. For example, `col1 == "A"`.

## Dataframes



## Submit progress

As you have gone through this tutorial, your progress has been kept track of in something called a "hash". I can use this to make sure you have executed all of the code chunks and answered all of the questions. I will use this to give you a completion grade for this assignment. 

If a Google Form window does not appear, you can access it [here](https://docs.google.com/forms/d/e/1FAIpQLScVRKB7vRsjD-Swc3RzSc01G-eCDZF4cYI3TI89dYOhTMdW2Q/viewform).

```{r context="server"}
learnrhash::encoder_logic(strip_output = T)
```

```{r encode, echo=FALSE}
learnrhash::encoder_ui(
  ui_before = div(strong("Submit your hash in the form below."), br(), br()),
  ui_after  = learnrhash::iframe_ui(
    src = "https://docs.google.com/forms/d/e/1FAIpQLScVRKB7vRsjD-Swc3RzSc01G-eCDZF4cYI3TI89dYOhTMdW2Q/viewform", 
    width="900px", height= "1000px"
  )
)
```


