---
title: "Capture-Recapture Explorer"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(learnrhash)
library(tidyverse)
#library(gradethis)
library(plotly)
#tutorial_options(
  # use gradethis for checking
#  exercise.checker = gradethis::grade_learnr
#  )
knitr::opts_chunk$set(echo = FALSE)
set.seed(113877)

```

These applets will help you explore estimators for abundance in a capture-recapture framework.

## Question 4 - Likelihood Plots

We want the value of $N$ that maximes the likelihood of this happening as our estimate. This would be a gnarly thing to maximize by hand. Let's investigate graphically.

For three different scenarios, use the hover feature to approximate the maximum of the likelihood curve. Write down the parameters you used and the values of the maximum points at those values of the parameters. Compare those values to those of the other two estimators we have seen so far (LP and MoM). Note: you can calculate these by plugging in the parameters you use in the applet.

Write your answer in your lab report.


```{r, echo=FALSE}
sliderInput("n_1", "tagged in first collection (n_1)", min = 10, max = 100, value = 10,step = 5)
sliderInput("m_2", "recaptured in second collection (m_2)", min = 0, max = 100, value = 5,step = 5)
sliderInput("n_2", "total in second collection (n_2)", min = 10, max = 100, value = 20,step = 5)

plotlyOutput("likPlot")
```

```{r, context="server"}
output$likPlot <- renderPlotly({
  N <- seq(0,100, by = 5)
  a <- choose(input$n_1, input$m_2)
  b <- choose(N-input$n_1, input$n_2-input$m_2)
  c <- choose(N, input$n_2)
  toPlot <- cbind.data.frame(N = N, lik = (a*b)/c)
 g <- ggplot(toPlot, aes(N, lik))+geom_point()+theme_minimal(base_size = 20)+xlab("N")+ylab("likelihood")
 ggplotly(g)
})
```



## Questions 6 and 7

The above estimators depends on three different parameters which are hard to vary simultaneously and assess their effect on the overall estimate. Let's examine some "profiles" of $N_{LP}$ aka look at different slices of this multi-dimensional space where we fix two parameters (as well as the true abundance) and vary the other one. 

**Question: When does this get close to the truth? Try a few combinations and note what you observe. Make sure to talk about the context of the parameters (i.e. treat $n_1$ as the number of animals seen on day one, not just as notation)**

**Question: So far we've been worried about underestimating since we will likely not detect some individuals. Try to find a situation where we overestimate the abundance by a lot. How likely is that situation to occur? Hint: consider the implied detection probability of the scenario.**

Write your answers in your lab report.


```{r,  echo=FALSE}
sliderInput("N2", "true abundance", min = 10, max = 100, value = 20,step = 5)
sliderInput("n_12", "number seen and tagged on day 1 (should be less than N)", min = 0, max = 100, value = 5,step = 5)
sliderInput("n_22", "number seen on day 2 (should be less than N)", min = 10, max = 100, value = 10,step = 5)

plotlyOutput("estPlot")
```

```{r,  context="server"}
output$estPlot <- renderPlotly({
  m_2 <- seq(0, input$N2, by = 1)
  est <- input$n_12*input$n_22/m_2
  toPlot <- cbind.data.frame(m_2 = m_2, est = est)
 g <- ggplot(toPlot, aes(m_2, est))+geom_point()+theme_minimal(base_size = 20)+xlab("number recaptured on day 2")+ylab("estimate")+geom_hline(yintercept = input$N2)+xlim(0, input$n_12)
 ggplotly(g)

})
```

```{r,  echo=FALSE}
sliderInput("N3", "true abundance", min = 10, max = 100, value = 20,step = 5)
sliderInput("n_13", "number seen and tagged on day 1 (should be less than N)", min = 0, max = 100, value = 10,step = 5)
sliderInput("m_22", "number recaptured on day 2 (should be less than n_1)", min = 0, max = 100, value = 5,step = 5)

plotlyOutput("estPlot2")
```

```{r,  context="server"}
output$estPlot2 <- renderPlotly({
  n_2 <- seq(0, input$N3, by = 1)
  est <- input$n_13*n_2/input$m_22
  toPlot <- cbind.data.frame(n_2 = n_2, est = est)

 g <- ggplot(toPlot, aes(n_2, est))+geom_point()+theme_minimal(base_size = 20)+xlab("number seen on day 2")+ylab("estimate")+geom_hline(yintercept = input$N3) 
 ggplotly(g)

})
```


## Question 8

The variance of the $N_{LP}$ estimator again is a function of three parameters, so we will want to look at some profiles to understand when the variance is small.

**Question: What situations minimize the variance? Try a few combinations and note what you observe. Make sure to talk about the context of the parameters**


Write your answer in your lab report.


```{r,  echo=FALSE}
sliderInput("N4", "true abundance", min = 10, max = 100, value = 40,step = 5)
sliderInput("n_14", "number seen and tagged on day 1 (should be less than N)", min = 0, max = 100, value = 15,step = 5)
sliderInput("n_24", "number seen on day 2 (should be less than N)", min = 10, max = 100, value = 20,step = 5)

plotlyOutput("varPlot")
```

```{r,  context="server"}
output$varPlot <- renderPlotly({
  m_2 <- seq(0, input$N4, by = 1)
  var <- ((input$n_14+1)*(input$n_24+1)*(input$n_14 - m_2)*(input$n_24-m_2))/((m_2+1)^2*(m_2+2))
  toPlot <- cbind.data.frame(m_2 = m_2, var = var)
 g <- ggplot(toPlot, aes(m_2, var))+geom_point()+theme_minimal(base_size = 20)+xlab("number recaptured on day 2")+ylab("variance of estimate")+xlim(0, input$n_14)
 ggplotly(g)

})
```


## Adjusting for the Small-Sample Bias

Nothing to write down here.


```{r,  echo=FALSE}
sliderInput("Na", "true abundance", min = 10, max = 100, value = 20,step = 5)
sliderInput("n_1a", "number seen and tagged on day 1 (should be less than N)", min = 0, max = 100, value = 5,step = 5)
sliderInput("n_2a", "number seen on day 2 (should be less than N)", min = 10, max = 100, value = 10,step = 5)

plotlyOutput("estPlot_adj")
```

```{r,  context="server"}
output$estPlot_adj <- renderPlotly({
  m_2 <- seq(0, input$Na, by = 1)
  est <- ((input$n_1a+1)*(input$n_2a+1))/(m_2+1)-1
  toPlot <- cbind.data.frame(m_2 = m_2, est = est)
 g <- ggplot(toPlot, aes(m_2, est))+geom_point()+theme_minimal(base_size = 20)+xlab("number recaptured on day 2")+ylab("estimate")+geom_hline(yintercept = input$Na)+xlim(0, input$n_1a)
 ggplotly(g)

})
```

